Distinct communities of snow algae bloom at different elevations

### 2019-10-17 doc/Casey1.Rmd
intro

Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo, by one estimate enough to substantially increase the rate of snow melt [@Ganey2017]. Loss of snow cover can contribute to drought [@Staudinger2014;@Barnhart2016] and exposes darker surfaces upon meltout, contributing to snow-albedo feedback [@Dery2007]. Thus, snow algae could play an important role in the cryosphere, yet little is known about what species of algae are present, or differences in community composition between snow algae blooms.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]. In the lab, cultures of *Chloromonas* are all morphologically indistinct green biflagellates, but field samples of this genus are dominated by cells with yellow or orange pigment and have thick cell walls with flanges or spikes [@Matsuzaki2019;@Prochazkova2019a]. Cells formerly defined by morphological similarity can be genetically distinct, for example cell types referred to as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->

<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
To date, genetic studies have identified three genera of algae that cause most snow algae blooms: *Chloromonas*, *Sanguina*, and *Chlainomonas*. Only *Chloromonas* contains cultured representatives, with _ named species on GenBank. *Chloromonas* are not known to produce the red pigment. *Sanguina* was recently designated, and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* are less studied, but appear in field samples as large (up to 40 $\mu$m diameter) quadriflagellate oval red cells. In spite of their striking morphological differences, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade.

@@

The habitat distribution of snow algae are not well established. Amplicon high-throughput sequencing (HTS) shows that red snow from sites across the Arctic are dominated by reads with high similarity to *Sanguina*, with low relative abundance of *Chloromonas polyptera*, *Chloromonas cf. nivalis*, and *Raphidonema* [@Lutz2016a]. @Ganey2018 suggests that the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to moister areas, although red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c]. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. In the Arctic, one green snow patch was dominated by *Microglena* [@Lutz2015]. Microscopy suggests differences in the snow algae community above and below treeline: two studies found red snow containing *Sanguina*-like cells to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a].

<!-- clear objectives section -->
 Whether snow algae blooms are typically dominated by a single species, or contain a diversity of species remains an open question, as high-resolution genetic comparison between communities is lacking.

Thus far, efforts to characterize snow algae communities by amplicon high-throughput sequencing (HTS) have been limited by ambiguous or nonexistent reference sequences and high 18S similarity between snow algae species [@Lutz2018]. To robustly compare snow algae community comparison side-by-side we used rbcL and 18S amplicon HTS and microscopy to survey 35 snow algae communities in our local mountains, to see what species were present, to compare community composition between samples, and see how they are distributed across the landscape.

>>>>

Snow algae are cold-adapted microalgae that bloom in melting snow [@Leya2013]. Red pigments produced by some species of snow algae can color vast areas of polar and alpine snowfields, which decreases snow surface albedo---by one estimate, enough to substantially increase the rate of snow melt [@Ganey2017]. Loss of snow cover can lead to snow drought [@Staudinger2014;@Barnhart2016], and upon meltout exposes darker surfaces below which contributes to snow-albedo feedback [@Dery2007]. Thus, snow algae could play an important role in cryosphere regulation, yet the species composition within snow algae blooms remains poorly studied. Snow algae are the primary producers in a microbiome that includes grazers and heterotrophic fungi and bacteria (ref), and can be considered ecosystem engineers through their impacts on snow melt and by supplying carbon to this poorly understood microbiome. Knowing the identity and community composition of the algae themselves is of primary importance.

Until recently, snow algae species were defined and identified on the basis of morphology. Morphology can be deceptive when identifying microalgae, as the same species can look completely different from one life stage to the next [@Shah2016;@Salome2019]; conversely, different species can look nearly identical. In the lab cultures of *Chloromonas* are green biflagellates, but in the field the same species contain yellow or orange pigment, and thick cell walls with flanges or spikes. Different species were thought to have diagnostic traits such as pigments or cell walls, but genetic studies show that morphological similarity may be misleading in this genus; for example, cell types formerly known as *Chloromonas cf. nivalis* are now split into multiple species [@Matsuzaki2019;@Prochazhova2019a].
<!-- based on this paragraph, I would not trust the morphology categories in my thing. Ill take em out anyway, but... chlainomonas is distinct, chloro distinct -->

<!-- get reader up to date on snow algae taxonomy... two out of three are poorly explored -->
To date, genetic studies have identified three main genera of algae that cause snow algae blooms: *Sanguina*, *Chlainomonas*, and *Chloromonas*. Neither of the first two have cultured representatives, but both produce the red pigment astaxanthin (refs). *Sanguina* was only recently designated and contains just two species, one of which *Sanguina nivaloides* is found in alpine and polar snowfields worldwide, and is likely synonymous with some of *Chlamydomonas cf. nivalis* [@Prochazkova2018c]. *Chlainomonas* also produce red pigment, but are physically much larger than *Sanguina* (up to 40 $\mu$m diameter). Two species are described on the basis of morphology, although analysis of the genetic differences between these species is lacking. Although *Chlainomonas* are morphologically distinct from all known *Chloromonas*, *rbc*L and 18S phylogeny places them within the *Chloromonas* clade. *Chloromonas* are highly diverse, with at least 13 named species [@Matsuzaki2019].

@@

Whether different species of snow algae are restricted to a particular microhabitat within the melting snowpack remains unresolved. Liquid water availability and nutrient limitation likely play a role in determining where snow algae blooms, as water and fertilizer application increased cell counts of *Chlamydomonas cf. nivalis* in experimental manipulation plots in Alaska [@Ganey2017]. Green snow blooms dominated by *Chloromonas* are reported from forested sites and shaded snowbanks [@Terashima2017;@Remias2013a], although some *Chloromonas* may be specially adapted for high light intensity in the alpine [@Prochazkova2019a]. Two studies using microscopy found that red snow containing *Chlamydomonas cf. nivalis* to be prevalent in the alpine zone above treeline, while green and orange snows caused by *Chloromonas cf. brevispina* and *Chloromonas cf. nivalis* were prevalent in the subalpine zone below treeline [@Hoham1979a;@Nedbalova2008a]. @Ganey2018 hypothesized the red pigment of snow algae may be a special adaptation to drier snow, while green snow is restricted to snow with higher moisture content; however red snow caused by *Chlainomonas* is only reported from waterlogged snow overlying alpine lakes [@Novis2008;@Prochazkova2018c].

Microscopic observation of snow algae blooms shows snow algae blooms can be either morphologically diverse [@Lutz2018] or highly similar [@Remias2013]. 18S amplicon high-throughput sequencing (HTS) shows that snow algae may contain more diversity than meets the eye: although red snow from sites across the Arctic were dominated by uncultured Chlamydomonadaceae, they also contained lower abundance of reads assigned to *Chloromonas* and *Raphidonema* [@Lutz2016a]. 18S is highly conserved and therefore suited to survey broadly across a variety of taxa (ref); however due to its high similarity between closely related species it is poorly suited to differentiating between closely related taxa. The *rbc*L plastid gene coding for the large subunit of RuBisCO is highly differentiated among photosynthetic organisms, and contains reference sequences for snow algae, making it an ideal complement to survey snow algae communities.

In this study, we surveyed snow algae communities from the mountains surrounding Vancouver, BC, Canada using light microscopy, 18S and *rbc*L amplicon HTS. Our aims were to: 1) see what species were present, 2) compare community composition between samples, 3) and see how they are distributed across the landscape.
